# ISIMIP Project: Download and Data Preprocessing

The **Inter-Sectoral Impact Model Intercomparison Project** develops and provides climate and socioeconomic forcing datasets with a spatial resolution of 1 degree, along with climate projections under SSP-RCP scenarios from 2015 to 2100.

The datasets from the ISIMIP3b protocol provide bias-corrected CMIP6 climate forcing for historical data and SSP1-RCP2.6, SSP3-RCP7.0, and SSP5-RCP8.5 conditions. The bias adjustment corrects the simulated data based on corrected ERA5 observational data (W5E5).

More information: <https://www.isimip.org/about/>

Try executing the chunks by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter*.

### Install or activate the required libraries

```{r}
#install.packages(c("utils", "ecmwfr", "gtools", "ncdf4", "ncdf4.helpers", "reshape2", "tidyverse", "sf"))

library(utils)
library(ecmwfr)
library(gtools)
library(ncdf4)
library(ncdf4.helpers)
library(reshape2)
library(tidyverse)
library(sf)
```

## 1. Set working directory

In the next step, you can choose the folder where the results will be stored.

```{r}
# Set the directory for data storage
user_wd <- readline(prompt = "Please enter your directory path: ")
user_wd <- gsub('"', '', user_wd); user_wd <- gsub('\\\\','/',user_wd)

while (!dir.exists(user_wd)) {
  print("Invalid directory. Please enter a valid one.")
  user_wd <- readline(prompt = "Please enter your directory path: ")
  user_wd <- gsub('"', '', user_wd); user_wd <- gsub('\\\\','/',user_wd)
}
print(paste("You entered a valid directory:", user_wd))
```

## 2. Enter the variable, time frame, and SSP climate change scenario of your interest

This code prompts the user to select a variable (precipitation, temperature, or evapotranspiration), a time frame (historical or future), and, if applicable, an SSP scenario (ssp126, ssp370, ssp585, or all).

### Scenarios

-   **historical**: GCM-based historical climate-related forcing, historical changes in CO2 and CH4 concentrations.
-   **ssp126**: GCM-based climate-related forcing according to SSP1-2.6.
-   **ssp370**: GCM-based climate-related forcing according to SSP3-7.0.
-   **ssp585**: GCM-based climate-related forcing according to SSP5-8.5.

```{r}
# Read csv file with links to NetCDF files
file <- paste0("/link_list.csv")
link_list <- read.csv(paste0(user_wd, file))

# Function to ask for valid input
ask_input <- function(prompt, valid_options) {
  repeat {
    input <- readline(prompt)
    if (input %in% valid_options) return(input)
    cat("Invalid input. Please choose from:", paste(valid_options, collapse = ", "), "\n")
  }
}

# Get valid input for variable and scenario
variable <- ask_input("Which variable are you interested in: Precipitation 'pr', Temperature 'tas', or Evapotranspiration 'potevap'? ", c("pr", "tas", "potevap"))
time_frame <- ask_input("Which time frame are you interested in: historical (1850-2014) or future (2015-2100)? ", c("historical", "future"))

# Set ssp_scenario to historical if time_frame is historical, otherwise ask for SSP scenario if time_frame is future
ssp_scenario <- if (time_frame == "historical") {
  "historical"  # Automatically set to historical if time frame is historical
} else {
  ask_input("Which SSP climate change scenario(s) are you interested in (ssp126, ssp370, ssp585, or all)? ", c("ssp126", "ssp370", "ssp585", "all"))
}

# Filter the data based on the user's input
if (ssp_scenario == "all") {
  # If 'all' is chosen, select any of the SSP scenarios
  filtered_df <- link_list[link_list$variable == variable & link_list$scenario %in% c("ssp126", "ssp370", "ssp585"), ]
} else {
  # If a specific SSP or historical scenario is chosen, filter for that scenario only
  filtered_df <- link_list[link_list$variable == variable & link_list$scenario == ssp_scenario, ]
}

# Output the user's choices
cat("You selected the variable:", variable, "\n")
cat("You selected the time frame:", time_frame, "\n")
cat("You selected the SSP scenario:", ssp_scenario, "\n")

# Create destination folder
temp_folder <- paste0("temp_", variable, "_", ssp_scenario)
temp_wd <- file.path(user_wd, temp_folder)

# Create the folder if it doesn't exist
if (!dir.exists(temp_wd)) {
  dir.create(temp_wd)
}

cat("Temp folder successfully created in working directory at:", user_wd, "\n")
```

## 3. Set paths to shapefile or csv with set of coordinates

In the next step, you can either select a shapefile representing the region (polygon) of interest or choose a CSV file containing the coordinates of interest.

```{r}
# Set the path to a. shapefile or b. CSV file with coordinates 
user_choice <- readline(prompt = "Please enter 'a' to input the location of your shapefile or 'b' to for CSV with coordinates: ")

if (tolower(user_choice) == "a") {
  # Read shapefile
  shp_path <- readline(prompt = "Please enter the path to your shapefile. Example: path/to/your/folder/polygon.shp :")
  shp_path <- gsub('"', '', shp_path); shp_path <- gsub('\\\\','/',shp_path)

  while (!file.exists(shp_path)) {
    print("Invalid file path. Please enter a valid one.")
    shp_path <- readline(prompt = "Please enter the path to your shapefile. Example: path/to/your/folder/polygon.shp :")
    shp_path <- gsub('"', '', shp_path); shp_path <- gsub('\\\\','/',shp_path)
  }

  shp <- st_read(shp_path)
  print(paste("You entered a valid path for the shapefile:", shp_path))

} else if (tolower(user_choice) == "b") {
  # Read CSV
  coord_path <- readline(prompt = "Please enter the path to your CSV with coordinates. Format: two columns latitude longitude. Example: path/to/your/folder/coordinates.csv :")
  coord_path <- gsub('"', '', coord_path); coord_path <- gsub('\\\\','/',coord_path)

  while (!file.exists(coord_path)) {
    print("Invalid file path. Please enter a valid one.")
    coord_path <- readline(prompt = "Please enter the path to your CSV with coordinates. 
                           Format: two columns latitude longitude. Example: path/to/your/folder/coordinates.csv :")
    coord_path <- gsub('"', '', coord_path); coord_path <- gsub('\\\\','/',coord_path)
  }

  coord_df <- read.csv(coord_path)
  print(paste("You entered a valid path for the CSV file:", coord_path))

} else {
  cat("Invalid choice. Please enter 'a' or 'b'.\n")
}
```

## 4. Download the dataset of interest

**Connect to the Server where the dataset is located**

The dataset will be downloaded for the assigned variable and years and stored in the pre-determined directory on your local computer.

```{r}
# Extract 'link' column from filtered data frame
links_list <- filtered_df$link

# Define the path for storing downloaded NetCDF files
nc_path <- file.path(temp_wd, "nc_files")

# Create the directory if it doesn't exist
if (!dir.exists(nc_path)) {
  dir.create(nc_path)
}

# Loop over the links to download the NetCDF files
for (link in links_list) {
  file_name <- basename(link)  # Extract file name from the link
  download_path <- file.path(nc_path, file_name)  # Set download path
  
  # Execute the curl command to download the file
  system(paste("curl -# -o", download_path, link))  
  
  # Pause for 5 seconds between downloads
  Sys.sleep(5)
  
  # Check if the file was downloaded successfully
  if (file.exists(download_path)) {
    cat("Download successful:", file_name, "\n")
  } else {
    cat("Download failed for:", file_name, "\n")
  }
  
  cat("\014")  # Clears the console for better visibility between requests
}

# Final confirmation message after all downloads are attempted
cat("Download complete. All files have been attempted and saved to:", nc_files, "\n")

```

## 5. Preprocessing of NetCDF dataset

If you are interested in preprocessing the data for a shapefile, please run the chunks i and ii from section 5.1. If you are interested in a CSV file, please run section 5.2.

After running the following chunks, the results will be stored in a folder called 'results' within your specified working directory.

### 5.1. CSV with coordinates

#### i. Average daily values for the list of coordinates of interest

```{r}

# Create folder to store output data
results_csv_path <- file.path(temp_wd, "results_csv")
if (!dir.exists(results_csv_path)) {
  dir.create(results_csv_path)
}

# Ensure id from set of coordinates
coord_df$id <- seq_len(nrow(coord_df))

# List all NetCDF files in the directory
nc_files <- list.files(path = file.path(nc_path), pattern = "\\.nc$", full.names = TRUE)

# Group files by climate forcing
grouped_nc_files <- split(nc_files, sapply(nc_files, function(file) {
  file_name <- basename(file)
  if (variable == "potevap") {
    climate_forcing <- sub("^watergap2-2e_([^_]+)_.*", "\\1", file_name)  # Custom extraction for 'potevap'
  } else {
    climate_forcing <- sub("^(.*?)_.*", "\\1", file_name)  # Default extraction for other variables
  }
  return(climate_forcing)
}))

# Loop over each climate forcing group
for (forcing_group in names(grouped_nc_files)) {
  
  files_to_process <- grouped_nc_files[[forcing_group]]
  
  # Loop over each coordinate set
  for (i in 1:nrow(coord_df)) {
    
    lat <- coord_df$lat[i]
    lon <- coord_df$lon[i]
    
    # Initialize empty data frame to store all extracted data
    extracted_data <- data.frame()
    
    # Loop through NetCDF files
    for (file in files_to_process) {
      
      file_name <- basename(file)
      
      # Extract scenario
      climate_scenario <- sub(".*_w5e5_(.*?)_.*", "\\1", file_name)
      
      # Open NetCDF file
      nc_file <- nc_open(file)
      variable_name <- names(nc_file$var)[1]  # Extract variable name
      
      # Extract metadata
      unit <- nc_file[["var"]][[variable_name]][["units"]]
      variable_description <- nc_file[["var"]][[variable_name]][["longname"]]
      
      # Extract time
      time_values <- nc_file$dim$time$vals
      time_units <- nc_file$dim$time$units
      reference_time <- as.POSIXct(gsub("days since ", "", time_units), format = "%Y-%m-%d %H:%M:%S")
      date_series <- as.Date(reference_time + (time_values + 1) * 86400)
      
      # Extract lat/lon indices
      lat_vals <- nc_file$dim$lat$vals
      lon_vals <- nc_file$dim$lon$vals
      closest_lat_index <- which.min(abs(lat_vals - lat))
      closest_lon_index <- which.min(abs(lon_vals - lon))
      
      # Extract data
      extracted_values <- ncvar_get(nc_file, variable_name, start = c(closest_lon_index, closest_lat_index, 1), 
                                    count = c(1, 1, -1))
      if (variable_name == "pr" | variable_name == "potevap") {
        extracted_values <- extracted_values * 86400  # Units conversion: mm day-1
        unit <- "mm day-1"
      }
      extracted_values <- round(as.vector(extracted_values), 4)
      
      nc_close(nc_file)
      
      # Create data frame and store scenario data
      temp_data <- data.frame(date = date_series, 
                              value = extracted_values, 
                              scenario = climate_scenario)
      
      # Append to results data frame
      extracted_data <- rbind(extracted_data, temp_data)
      
      print(paste0("Results extracted: ", file_name))
      
    }
    
    # Reshape data to have separate columns for SSP scenarios
    final_data <- spread(extracted_data, key = scenario, value = value)
    
    # Rename columns for clarity
    colnames(final_data) <- gsub("ssp", "SSP_", colnames(final_data))
    
    # Construct output file path
    output_filepath <- file.path(results_csv_path, 
                                 paste0("id_", coord_df$id[i], "_climate_forcing_", 
                                        forcing_group, "_daily_", variable_name, ".txt"))
    
    # Open connection for writing
    con <- file(output_filepath, 'w')
    
    # Write metadata
    cat(file = con, sprintf("# Repository    :\t%s\n", "https://data.isimip.org/search/tree/ISIMIP3b/"))
    cat(file = con, sprintf("# Climate forcing           :\t%s\n", forcing_group))
    cat(file = con, sprintf("# Variable      :\t%s\n", variable_description))
    cat(file = con, sprintf("# Units         :\t%s\n", unit))
    cat(file = con, sprintf("# Latitude      :\t%s\n", lat))
    cat(file = con, sprintf("# Longitude     :\t%s\n\n", lon))
    
    # Write final data table (date, SSP_126, SSP_370, SSP_585)
    write.table(final_data, file = con, sep = '\t', dec = '.', row.names = FALSE, quote = FALSE, na = "NA")
    
    close(con)
    
    print(paste0("Results saved: ", output_filepath))
    
    cat("\014")  # Clears the console for better visibility between requests
    
  }
}
```
